---
title: "Ben Harrison global metabolomics data negative mode - reanalyze pid1652"
author: Lu Wang
output:
   BiocStyle::html_document:
       toc_float: true
       css: style.css
bibliography: Statistical_analysis.bib
---
  
  
  
```{r setup, include=FALSE}
  library(knitr)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(BiocStyle)
  library(limma)
  library(edgeR)
  library(threejs)
  library(htmlwidgets)
  library(xtable)
  library(ReportingTools)
  library(plotly)
  library(openxlsx)
  library(DT)
  library(affycoretools)
  library(Glimma)
  #library(sva)
 library(ComplexHeatmap)

```


```{r parameter, include=FALSE}
opts_chunk$set(echo=FALSE, fig_path="plots/tmp")
# set parameters
fdrcutoff=0.1
missingCutoff =5
datamode = "negative"
removecol="posLabel"
datapath="../../../Metabolomics2016/Newdatafiles/pid1652_Negative_sample/XCMS-Negdata-93samples-Promislow.xlsx"
```


```{r main, include=FALSE}
setwd(paste0("/data5/analyses/Daniel_Promislow/Ben_Harrison_pid2005/",datamode,"_mode/bioconductor"))
# re-analyze pid1652 data, use new pipeline and need to adjust for the batch (block) as confounder.
# read in sample information.
samps= read.csv("../../../Metabolomics2016/OlddataFiles/sample.info.noqc.csv",stringsAsFactors = F)
samps= select(samps, -removecol)
samps$treatment <- as.factor(ifelse(samps$treatment =="B-", "Ctr", "H2O2"))
samps$trait <- factor(ifelse(samps$lifespan >= 80, "resistant", "sensitive"),
                      levels = c("sensitive","resistant"))

colnames(samps)[1] <- "Sample.ID"
# re-order by the run order
samps = samps[order(samps$runOrder),]
# Some summary tables
table(samps$treatment)
table(samps$trait)
table(samps$treatment,samps$trait)
# 16 lines
samps$line=gsub("-","_", samps$line)
length(unique(samps$line))
samps$line = as.factor(samps$line)
# 3 BATCH!!!
table(samps$line,samps$block)
samps$block = as.factor(samps$block)
samps$comb = factor(paste(samps$trait,samps$treatment, sep="_"),
                       levels=c("sensitive_Ctr", "sensitive_H2O2", "resistant_Ctr", "resistant_H2O2"))
#-----------------------------------------------------
raw.data = read.xlsx(datapath)
colnames(raw.data) = gsub("name","metabolites", colnames(raw.data))
any(duplicated(raw.data$metabolites)) #FALSE
sum(! is.na(raw.data$CV))
row.names(raw.data) = raw.data$metabolites
colnames(raw.data) <- gsub("\\.mzdata", "", colnames(raw.data))
colnames(raw.data) <- gsub("_", "", colnames(raw.data))
# one sample named "neg46.2"; not sure what dose'.2' means;
(colnames(raw.data) <- gsub("\\.2","", colnames(raw.data)))
stopifnot(all(samps$Sample.ID %in% colnames(raw.data)))
#features info
feature_info = select(raw.data, featureidx:postHoc)
# do log2 transformation, replace zero with NA.
cleanup <- function(input, selected_column){
  dat= dplyr::select(input, starts_with(selected_column))

  print(colnames(dat))
  dat[dat==0] <-NA
  dat = as.matrix(log2(dat))

  print(paste0("missing%=",round(mean(is.na(dat))*100)))
  print(apply(na.omit(dat),2, range))
  
  outlist = list("dat" = dat)
  return(outlist)
}
#----------------------------------
# QC sample
sample.qc =  cleanup(raw.data, "QC")$dat
o=order(as.integer(gsub("QC(neg|pos)","",colnames(sample.qc))))
# re-order as the runorder
sample.qc = sample.qc[,o]
colnames(sample.qc)
# create a CV% for QC sample.
CV= data.frame(metabolites = row.names(sample.qc),
               CV=(apply(sample.qc,1,function(x) 
                 sd(x, na.rm = T))/rowMeans(sample.qc,na.rm = T))*100,
                                stringsAsFactors = F)
#----------------------------------
substr(datamode,1,3)
raw.matrix =  cleanup(raw.data,substr(datamode,1,3))$dat
raw.matrix = raw.matrix[, match(samps$Sample.ID,colnames(raw.matrix))]
stopifnot(all.equal(colnames(raw.matrix), samps$Sample.ID)) #TRUE
samps$percent_missings = round(colMeans(is.na(raw.matrix))*100,2)
stopifnot(all.equal(row.names(sample.qc), row.names(raw.matrix)))
#----------------------------------
# EDA
if(!dir.exists("plots")) dir.create("plots")
mycol <- c("#00B2EE", "#FFC125", "#A8A8A8")

pdf(paste0("./plots/Density_plot_",datamode,".pdf"))
plot(density(raw.matrix[,1], na.rm=TRUE), col=mycol[1], lty=2,   main = "Desnsity plot") 
apply(raw.matrix[,2:ncol(raw.matrix)], 2, function(x) lines(density(x, na.rm=TRUE), col=mycol[1], lty=2))
apply(sample.qc, 2, function(x) lines(density(x, na.rm=TRUE), col=mycol[2]))
legend("topright", col=mycol, legend = c("Samples", "QC"), lty =1)
dev.off()
#-----------------------------------
# BOX plot
pdf(paste0("plots/boxplots_",datamode,".pdf"), width = 12)
boxplot(raw.matrix,las=2,col=mycol[1],ylab="log2 abundance", main="Sample")
boxplot(sample.qc,las=2,col=mycol[2],ylab="log2 abundance", main="QC")
dev.off()
#---------------------------------------------
pdf(paste0("./plots/sample_hist_",datamode,".pdf"))
par(mfrow = c(2,2))
hist(colSums(raw.matrix,na.rm = T),breaks = 40, col="light blue",  main = "Sum of log2(abundance) by sample")
hist(colMeans(raw.matrix,na.rm = T),breaks = 40, col="yellow",main = "Mean of log2(abundance) by sample ")
hist(rowSums(raw.matrix,na.rm = T),breaks = 40, col="light blue", main = "Sum of log2(abundance) by feature ")
hist(rowMeans(raw.matrix,na.rm = T),breaks = 40,col="yellow", main = "Mean of log2(abundance) by feature ")
hist(colMeans(is.na(raw.matrix)),breaks = 40, 
     main = "Percentage of missings by sample")
hist(rowMeans(is.na(raw.matrix)),breaks = 40,
     main = "Percentage of missings by feature")
dev.off()
#-----------------------------------------
# Missing data
# overall percent of missing data 
mean(is.na(raw.matrix))*100 #2%
# percent of features have < 5% missing data
mean(rowMeans(is.na(raw.matrix))*100 < 5)*100
# percent missing per sample
colMeans(is.na(raw.matrix))*100
#-----------------------------------
# missing by feature, across all samples  
missing_by_feature = data.frame(metabolites = row.names(raw.matrix),
                                mean_log2abundance_noNA = rowMeans(raw.matrix,na.rm = T),
                                percent_missing = rowMeans(is.na(raw.matrix))*100,
                                stringsAsFactors = F)

missing_by_feature = dplyr::left_join(missing_by_feature,CV,by="metabolites")
# plot the percent_missing over mean_log2abundance
pdf(paste0("./plots/percent_missing_meanabundance_",datamode,".pdf"))
plot(missing_by_feature$mean_log2abundance_noNA, missing_by_feature$percent_missing, 
     pch=19, col="light blue", xlab="mean log2(abundance)", ylab="Percent of missingness", 
     main = datamode)
points(missing_by_feature$mean_log2abundance_noNA[missing_by_feature$percent_missing>=missingCutoff],
       missing_by_feature$percent_missing[missing_by_feature$percent_missing>=missingCutoff], pch=19, col="orange")
legend("topright",
       legend=c(paste0("<",missingCutoff,"% missingness"), 
                paste0(">=",missingCutoff,"% missingness")),pch=19, cex=0.8,
       col=c("light blue", "orange"))
dev.off()
#----------------------------------------------
table(missing_by_feature$percent_missing)
samps %>% group_by(comb) %>% 
  summarise(mean(percent_missings))

ms.plot = ggplot(samps, aes(x= reorder(Sample.ID, percent_missings), y= percent_missings,       
                 fill= comb)) +
                 geom_bar(stat="identity") + 
                 labs(title = paste0("Sample missing data "), 
                 x = "",  y= "Percent of missingness", fill="") +
                 theme(legend.position = "top",
                        axis.text.x=element_text(angle=90, hjust=1)) 

pdf("./plots/Percent_missing_sample.pdf")
ms.plot
dev.off()
#----------------------------------------------
# 3D-pca
source("pca_set.R",local = T)
pca_set(input = raw.matrix,sample_df = samps,title = paste0("Raw_data_",datamode))
#----------------------------------------------
# median normalization for two tissue separately.
apply(raw.matrix,2, function(x) median(x,na.rm = T))
(median.global <- median(raw.matrix, na.rm = TRUE))
dat.matrix <- apply(raw.matrix, 2, function(x) x - median(x, na.rm = TRUE) + median.global)

pdf(paste0("./plots/normalization_boxplots_",datamode,".pdf"))
boxplot(raw.matrix,las=2,col=mycol[1],ylab="log2(abundance)", main="Before normalization", cex.axis =0.6)
boxplot(dat.matrix,las=2,col=mycol[2],ylab="log2(abundance)", main="After normalization", cex.axis = 0.6)
dev.off()

# select features with missingness <5
select_feature = missing_by_feature[missing_by_feature$percent_missing <missingCutoff & missing_by_feature$CV<30,]$metabolites

dat.matrix.filter = dat.matrix[row.names(dat.matrix) %in% select_feature,]
dim(dat.matrix.filter) 
mean(is.na(dat.matrix.filter))*100
range(apply(dat.matrix.filter, 2, function(x) mean(is.na(x))*100))
#limma::plotDensities(dat.matrix.filter,legend = F)
# use KNN imputation
impute.matrix = impute::impute.knn(dat.matrix.filter)$data
pca_set(input = impute.matrix,sample_df = samps,title = paste0("Imputed_",datamode))
#limma::plotDensities(impute.matrix,legend = F)
#---------------------------------------------------------
stopifnot(all.equal(colnames(impute.matrix),samps$Sample.ID))
#----------------------------------------------
# check drifting
zz.long = data.frame(raw.matrix) %>%
  mutate(metabolites = row.names(.))%>%
  gather(., Sample.ID, log2abundance,-metabolites) %>%
  left_join(., samps,by="Sample.ID")

pdf(paste0("plots/",datamode,"_metabolite_dotplot_before_normalization.pdf"),height = 180,width = 18)
ggplot(zz.long[zz.long$metabolites %in% select_feature[1:200],],aes(x=runOrder, y =log2abundance,color=comb))+
  geom_point()+
  facet_wrap(~metabolites, ncol=2)+
  theme(legend.position = "top", legend.direction = "horizontal")
dev.off()
#---------------------------------------------
source("pca_set2.R",local = T)

# check what other covariates need to be adjusted besides batch.
noBatch = removeBatchEffect(impute.matrix,design = model.matrix(~comb,samps),
                        batch = samps$block)
pca_set(noBatch, samps, title= paste0("noBatch_",datamode))

pca_set2(noBatch, samps, title= paste0("noBatch_2",datamode))

for (i in 1:length(unique(samps$comb))){
  subset = samps[samps$comb ==levels(samps$comb)[i],]
  pca_set2(noBatch[, match(subset$Sample.ID,colnames(noBatch))], subset, title=      
                 paste0("noBatch_2",datamode,levels(samps$comb)[i]))
}
source("pca_set3.R",local = T)
for (i in 1:length(unique(samps$comb))){
  subset = samps[samps$comb ==levels(samps$comb)[i],]
  pca_set3(noBatch[, match(subset$Sample.ID,colnames(noBatch))], subset, title= paste0("noBatch_3",datamode,levels(samps$comb)[i]))
}

noBatch_pca= pca_set2(noBatch, samps, title=paste0("noBatch_",datamode))$pca_df
summary(lm(noBatch_pca$PC3 ~samps$sampleWeight+samps$comb))# need to be adjusted
summary(lm(noBatch_pca$PC3 ~samps$line.weight+samps$comb))
summary(lm(noBatch_pca$PC3 ~samps$sampleWeight+samps$line.weight+samps$comb))# 
#---------------------------------------------------------
# use limma pipeline
#----------------------------------------
levels(samps$comb)
design <- model.matrix(~0 + comb + block + sampleWeight, data = samps)
colnames(design) = gsub("comb","", colnames(design))
colnames(design)
contrast = makeContrasts(resistant_H2O2 - resistant_Ctr,
                         sensitive_H2O2 - sensitive_Ctr,
                         resistant_Ctr - sensitive_Ctr,
                         resistant_H2O2 - sensitive_H2O2,
                         (resistant_H2O2 - resistant_Ctr) - (sensitive_H2O2 - sensitive_Ctr),
                         levels = design)
colnames(contrast) = gsub(" - ","_vs_", colnames(contrast))

fit <- lmFit(impute.matrix, design = design)
fit <- contrasts.fit(fit, contrast)
fit <- eBayes(fit,robust = T,trend = T)
plotSA(fit)
out <- lapply(1:ncol(contrast), function(x) topTable(fit, coef = x, n= Inf, adjust = "BH",p.value=fdrcutoff, sort.by = "p")[,-6])
sapply(out,nrow) 
#-------------------
out2 <- lapply(1:ncol(contrast), function(x) topTable(fit, coef = x, n= Inf, adjust = "BH",p.value=1, sort.by = "none")[,-6])
pdf("plots/hist_pvalue.pdf")
par(mfrow = c(2,2))
for (i in 1:length(out2)){
   hist(out2[[i]]$P.Value, main=paste0("out2_",colnames(contrast)[i]))
}
dev.off()
#---------------------------------------------
sapply(out2,nrow)
names(out2) = colnames(contrast)
outputname = colnames(contrast)
outputname[ncol(contrast)] <- "Interaction"
#---------------------------------------------
# remove batch and sample weight for plotting purpose only.
noCV = removeBatchEffect(impute.matrix,design = model.matrix(~comb,samps),
                        batch = samps$block, covariates = samps$sampleWeight)

pca_set(noCV, samps, title= paste0("noCV_",datamode))
for (i in 1:length(unique(samps$treatment))){
  subset = samps[samps$treatment ==levels(samps$treatment)[i],]
  pca_set2(noCV[, match(subset$Sample.ID,colnames(noCV))], subset, 
           title= paste0("noCV_2",datamode,levels(samps$treatment)[i]))
}

noCV_pca= pca_set(noCV, samps, title=paste0("noCV_",datamode))$pca_df
pve=pca_set(noCV, samps, title=paste0("noCV_",datamode))$pve
#---------------------------------------------
if(!file.exists("glimma_plots")) dir.create("glimma_plots")
for (i in 1:length(out2)){
  out2[[i]]$metabolites= row.names(out2[[i]])
  out2[[i]]= select(out2[[i]], metabolites, everything())
 
  status = ifelse((out2[[i]]$adj.P.Val<fdrcutoff & out2[[i]]$logFC>0),1,0)
  status = ifelse((out2[[i]]$adj.P.Val<fdrcutoff & out2[[i]]$logFC<0),-1,status)
  nameregrx = gsub("_vs_", "|",colnames(contrast)[i])
  stopifnot(all.equal(names(fit$coefficients[,i]), out2[[i]]$metabolites))
  glMDPlot(fit[,i], 
           counts =noCV[,grepl(nameregrx, samps$comb)], 
           status = status,
           anno = as.data.frame(out2[[i]]$metabolites), 
           samples = samps$Sample.ID[grepl(nameregrx, samps$comb)],
           main = outputname[i], 
           groups = samps$comb[grepl(nameregrx, samps$comb)],
           transform=FALSE,
           side.main="metabolites",
           display.columns= "metabolites",
           xlab="Average log2 abundance",
           side.ylab = "log2 abundance",
           html = paste0(outputname[i],"_MDplot"),
           launch = FALSE, folder = "glimma_plots")
}

MDplots = hwriter::hwrite(colnames(contrast),
               link = paste0("./glimma_plots/",outputname,"_MDplot.html"), table=FALSE)
#----------------------------------
if(!file.exists("spreadsheets")) dir.create("spreadsheets")
for (i in 1:ncol(contrast)){
     if (nrow(out[[i]])>0){
  colnames(out[[i]]) <- gsub("adj.P.Val", "FDR",colnames(out[[i]]))
  out[[i]]$metabolites = row.names(out[[i]])
  out[[i]]= dplyr::left_join(out[[i]],    
                            feature_info[,c("metabolites","mzmed","rtmed")],by="metabolites")
  out[[i]]= select(out[[i]], metabolites, everything())
  write.xlsx(out[[i]], paste0("spreadsheets/",datamode,"_",outputname[i],".xlsx"), row.names=F)
     }
}
#------------------------------------------
# heatmap for interaction
source("heatmaps_set.R",local = T)
if (nrow(out[[ncol(contrast)]])>0){
    heatmaps_set(input = noCV,selected_metabolite = out[[ncol(contrast)]]$metabolites,
             selected_sample = samps,title = paste0(datamode,"_treatment_trait_interactions"))
}
#---------------------------------------------
# ggplot2's stat_ellipse() is using car::dataEllipse(), which draws data ellipse not confidence ellipse.
#github.com/tidyverse/ggplot2/issues/2776
mycol2=c("#87CEFA", "#104E8B","#EEC900", "#CD6600")
tiff(paste0("plots/",datamode,"_noCV_2DPCA_Confidence95_ellipse.tiff"), units="in",width = 8, height=8,res =600, compression ="lzw")
ggpubr::ggscatter(data=noCV_pca,x="PC1",y ="PC2", 
                  xlab = paste0("PC1 (", pve[1],"%)"),
                  ylab = paste0("PC2 (", pve[2],"%)"), color="comb",
                  palette=mycol2,ellipse=TRUE, mean.point=FALSE,star.plot = TRUE,
                  ellipse.type = "confidence", ellipse.level = 0.95,show.legend.text=F,
                  ellipse.alpha=0.1)+
  labs(title = paste0("PCA plot on removed Batch & Sampleweight",datamode,"data - 95% confidence Ellipse"), colour = NULL,fill=NULL)+
  theme(axis.text = element_text(size = 12, face = "bold"))
dev.off()
#---------------------------------------------
tiff(paste0("plots/",datamode,"_noCV_2DPCA_Data_ellipse.tiff"), units="in",     
     width = 8, height=8,res =600, compression ="lzw")
ggplot(data= noCV_pca, aes(x=PC1, y=PC2, color = comb))+
  geom_point()+
  scale_colour_manual(values=mycol2) +  
  stat_ellipse(type = "t", level = 0.5)+
  labs(title = paste0("PCA plot on removed Batch & Sampleweight",datamode,"data - 0.5data Ellipse"),
       x = paste0("PC1 (", pve[1],"%)"),
       y = paste0("PC2 (", pve[2],"%)"),
       colour= NULL)+
  ggpubr::theme_pubr()+
  ggpubr::labs_pubr()
dev.off()
#---------------------------------------------
###Venn
if(!file.exists("venns")) dir.create("venns")
collist = list(c(1,2),c(3,4))

venns <- makeVenn(fit, contrast= contrast[1:4,],
                  design= design[,1:4], collist = collist,
                  p.value=fdrcutoff, method="both", affy=FALSE,
                  probecol="metabolites")

vennlink = vennPage(venns, pagename = "venn", shift.title = T,
                    pagetitle = paste0("Venn Diagram ", datamode))
#------------------------------------------
# output ipathway sheet.
names(out2) = outputname
all = do.call(cbind, out2)
colnames(all) = gsub("adj.P.Val|adj.P.value", "FDR", colnames(all))
colnames(all)[1] = sapply(strsplit(colnames(all)[1],"\\."),"[",2)
all = select(all, -ends_with(".AveExpr"))
all = select(all, -ends_with(".metabolites"))
all=dplyr::left_join(all,select(feature_info,  metabolites,mzmed, rtmed),by= "metabolites")
all.equal(all$metabolites,row.names(noCV))
all = cbind(all,noCV)
write.xlsx(all, paste0("Ben_Harrison_",datamode,"_analysis_all.xlsx"), row.names=F)
#---------------------------------------------
if(!file.exists("mummichog_input")) dir.create("mummichog_input")
#output for mummichog
out3=out2
for (i in 1:length(out3)){
  out3[[i]]= dplyr::left_join(out3[[i]], feature_info,by="metabolites") %>%
            select(., mzmed, rtmed, adj.P.Val,t)
  print(colnames(out3[[i]]) )

   if(nrow(out[[i]]) >0) 
     stopifnot(sum(out3[[i]]$adj.P.Val<fdrcutoff)==nrow(out[[i]]))
     write.table(out3[[i]],paste0("mummichog_input/",datamode,"_",outputname[i],".txt"),
                 sep="\t",quote = F,row.names = F)
}  
#------------------------------------------
save.image(paste0("Ben_Harrison_pid2005_",datamode,".Rdata"))
#-------------------------------------------
if(!file.exists("mummichog_output")) dir.create("mummichog_output")
setwd("mummichog_output")
(mummichog_inputfiles = list.files("../mummichog_input",full.names = T))
(mummichog_OUTfiles = list.files("../mummichog_input",full.names = F))
mummichog_OUTfiles = gsub(".txt","",mummichog_OUTfiles)

#Single file format (all features included)
for (i in seq_along(mummichog_inputfiles)){
cat(paste0("~/miniconda2/envs/mummichog/bin/mummichog1 -f ",mummichog_inputfiles[i],
           " -n fly",
           " -o ",mummichog_OUTfiles[i],
           " -m ",datamode, # mode
           " -c ",fdrcutoff, #FDR cutoff,
           " -p 100", "\n"#num of permutation
           ))
}
#-------------------------------------------
setwd("..")
```





```{r, echo = FALSE, warning=FALSE, message = FALSE, results = "asis"}

BiocStyle::markdown()

```


# Introduction

**This is a re-analysis of pid1652 `r datamode` data by considering the batch effect.**

In this global `r datamode` mode metabolomics data, there are a total of `r nrow(samps)` Drosophila samples. The samples are from `r length(unique(samps$line))` genetic fly lines and two treatment conditions (control or $H_2O_2$). The lifespan (trait) which was measured in the $H_2O_2$ treated group is dichotomized into two groups, where >=80 hr is designated as resistant and < 80 hr as sensitive group. Samples were prepared in 3 batches (block column). The table below shows the sample information.

The objectives of this project are:


  + To detect metabolites that have a significantly different abundance between treatment conditions within each trait group (e.g., $H_2O_2$ vs. control in resistant or sensitive flies).
  
  + To detect metabolites that have a significantly different abundance between traits within each treatment condition (e.g. resistant vs. sensitive in control or $H_2O_2$) .

  + To detect metabolites that have a significant treatment and trait interaction.
  
The table below shows the sample information. 

```{r, echo = FALSE, results = "asis"}
DT::datatable(samps, rownames= FALSE, caption =  "Samples used in this study (sorted by run order).")%>% 
  formatRound(columns= c(9,10,11,17),1)


kable(table(samps$treatment,samps$trait),
      caption = "The number of samples stratifided by treatment and trait groups")

kable(table(samps$line,samps$treatment),
      caption = "The number of samples stratifided by line and treatment")

kable(table(samps$comb,samps$block),
      caption = "The number of samples stratifided by group and batches")
```

The boxplot below shows the line weight (top) and sample weight (bottom) distribution across groups, where each dot represents a sample.  Notice that line weight is confounded wtih trait, such that resistant lines have higher weight than the sensitive lines.

```{r, echo = FALSE, results = "asis"}

ggplot(samps, aes(x=trait, y = line.weight, fill = trait)) +
  geom_boxplot(outlier.shape =  NA) +
  geom_point(position= position_dodge2(width = 0.2), alpha=0.3)+ 
  labs(x="", fill=NULL)+ theme(legend.position = "none")


ggplot(samps, aes(x=comb, y = sampleWeight, fill = comb)) +
  geom_boxplot(outlier.shape =  NA) +
  geom_point(position= position_dodge2(width = 0.2), alpha=0.3)+ 
  labs(x="", fill=NULL)+ theme(legend.position = "none")
```


# Exploratory data analysis (EDA) and Data preprocessing.

A useful way to check the quality of our data is a principal components analysis (PCA) plot. PCA is a method to take high dimensional data and reduce it to only a few dimensions in order to visualize how similar samples of the same type are. You can click on the legend to select or deselect the group, and if you hover your cursor over a given symbol, it will show you the sample information.

Below shows a series of PCA plots that are coded by different variables using the $log_2$ transformed raw data. PC1 separates the trait (e.g. resistant vs. sensitive group). PC2 separates the batches (block column). 


```{r, echo = FALSE, message=FALSE,warning=F}
pca_set(input = raw.matrix,sample_df = samps,title = paste0("Raw_data_",datamode))$pca.3d
pca_set(input = raw.matrix,sample_df = samps,title = paste0("Raw_data_",datamode))$pca.batch
pca_set(input = raw.matrix,sample_df = samps,title = paste0("Raw_data_",datamode))$pca.line
```

## Normalization

To remove systematic variation, we normalized the data by using a median normalization where we adjust the data so all samples have the same median value.

```{r, echo = FALSE, message=FALSE, fig.height=8, fig.width=12}

boxplot(raw.matrix,las=2,col=mycol[1],ylab="log2(abundance)", main="Before normalization", cex.axis =0.6)
boxplot(dat.matrix,las=2,col=mycol[2],ylab="log2(abundance)", main="After normalization", cex.axis = 0.6)
```


## Missing value imputation

There are a total of `r nrow(raw.data)` metabolites in the `r datamode` mode data, and `r round(mean(missing_by_feature$percent_missing[missing_by_feature$percent_missing!=100]),1)`% of the data are missing. There are `r sum(missing_by_feature$percent_missing==0)` metabolites with complete data (no missing values). The figure below shows the percent of missingness across samples, and the average percent missingness is about the same across groups.

```{r, echo = FALSE, message=FALSE, warning=FALSE}
ms.plot
tb2= samps %>% group_by(comb) %>%
  summarise(round(mean(percent_missings),1))
colnames(tb2) <- c("group","missingness%")
kable(tb2,caption = "Average percent missingness in each group.")
```

The figure below shows the percent missingness versus the mean $log_2$-abundance for each metabolite.


```{r, echo = FALSE, message=FALSE, fig.height=8}

plot(missing_by_feature$mean_log2abundance_noNA, missing_by_feature$percent_missing, 
     pch=19, col="light blue", xlab="mean log2(abundance)", ylab="Percent of missingness", 
     main =datamode)
points(missing_by_feature$mean_log2abundance_noNA[missing_by_feature$percent_missing>=missingCutoff],
       missing_by_feature$percent_missing[missing_by_feature$percent_missing>=missingCutoff], pch=19, col="orange")
legend("topright",
       legend=c(paste0("<",missingCutoff,"% missingness"), 
                paste0(">=",missingCutoff,"% missingness")),pch=19, cex=0.8,
       col=c("light blue", "orange"))

```


Missing values are known to be a problematic issue for mass-spectrometry (MS)-based data. Previous studies have summarized the three missing value mechanisms in terms of MS-based data [@Lazar:2016;@Wei:2018]: 
**Missing Completely At Random (MCAR)**, which corresponds to the combination and propagation of multiple minor errors or stochastic fluctuations. Each missing value cannot be directly explained by the nature of the feature or by its measured intensity. **Missing At Random (MAR)** assumes the possibility of a variable being missing is determined by other observed variables, e.g., inaccurate peak detection. MCAR is often combined with MAR when selecting imputation methods for MS analysis. MCAR/MAR widely occurres in the untargeted metabolomics data [@Wei:2018].

All the metabolites with >=`r missingCutoff`% missingness and with >= 30% coefficient of variation (CV) in QC samples were excluded, and a total of `r length(select_feature)` metabolites were included in the imputation step (roughly  `r round(mean(is.na(dat.matrix.filter)*100),2)`% of the data has missing values after filtering). We used the K-nearest neighbors (KNN) imputation method, which has shown good performance for imputing MCAR/MAR missing data [@Lazar:2016]. It is implemented in the R impute [@impute:2001] package. 

## More PCA plots

Below is a series of PCA plots after removing batch as a covariate on pre-processed data. Please notice that we will include batch as a covariate in the model directly. This step is just to check if there are any other confounders that need to be considered.

Overall, PC1 separates the resistant vs. sensitive groups, and to some extent separates the control and $H_2O_2$ treated flies in the sensitive group but not in the resistant group. PC1 & PC3 also may be related to sample weight. 


```{r, echo=FALSE, message= FALSE,warning=FALSE}
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.3d
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.batch
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.order
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.linewg
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.sampleWeight
pca_set(noBatch,sample_df =samps,title= paste0("noBatch_",datamode))$pca.sampmiss
```


# Model fit and results

We fit a linear model to the normalized and imputed data to detect the group differences using the Bioconductor limma package [@Ritchie:2015], while adjusting for the batch effect (e.g. block) and sample weight as covariates in our model. The limma package uses empirical Bayes moderated statistics, which improves power by 'borrowing strength' between metabolites in order to moderate the residual variance [@Smyth:2004]. We used the Benjamini-Hochberg multiple testing method to control the false discovery rate (FDR) and selected metabolites at an FDR of `r fdrcutoff*100`%. The last comparison is to test the treatment and trait interaction.

```{r, echo =FALSE, results='asis'}

sc <- function(x) x

d.f <- data.frame( Index = 1:ncol(contrast),
                   Comparisons = MDplots, 
                  "Number significant" = sapply(out, nrow),
                  check.names = FALSE)
print(xtable(d.f, caption = paste0("Number of significant metabolites at FDR < ",fdrcutoff)),
      type="html", include.rownames= FALSE, sanitize.text.function = sc)

```

The links in the above table will bring up an interactive HTML page that will assist you in exploring the results. The HTML page includes an interactive mean-difference (MD) plot (top left) where metabolite-wise $log_2$-fold changes (logFCs) are plotted on the vertical axis, against mean abundance values on the horizontal axis. Significantly differentially abundant metabolites are highlighted in red if they are increased and blue if they are decreased. If you hover over any of the points in the plot to the left, the $log_2$-abundance is displayed for that metabolite at the top right. A table of metabolite information is displayed at the bottom, and if you click a metabolite in the table, it will be highlighted on the top left mean-difference (MD) plot, and the top right will show the abundance levels of that metabolite in each of the samples. If you hover over each dot in the plot, it will show you more information. Please note that the table includes all of the metabolites ordered by the Adj.Pvalue (smallest to largest). 

* Average log2 abundance, the average log2-abundance of that metabolite across all samples.
* logFC, log2 fold-change of abundance between conditions being tested.
* Adj.Pvalue, FDR (using Benjamini-Hochberg procedure)


We also output the results table with information of all the individual metabolites regardless of the FDR cutoff in an Excel spreadsheet named `r paste0("Ben_Harrison_",datamode,"_analysis_all.xlsx")`, where metabolites are in rows and columns shows the logFC ($log_2$ fold-change), P.value (raw P.value) and adj.P.Val (FDR using Benjamini-Hochberg procedure) for each of the comparisons listed in the above table. 

## Venn Diagram

In order to identify common and unique metabolites between comparisons, we generated a list of two-way Venn diagrams. If you click the [hyperlink](venns/venn.html), it will load a page with Venn diagrams for each of the analysis that shows the number of metabolites changed among these comparisons. If you click on the counts in any cell of the Venn diagram, it will show you a table with metabolites in that cell. Please note that the p.value columns in that table are FDR.

## Heatmap

In order to visualize how resistant and sensitive group respond to treatment, we generated a heatmap for the metabolites that have a significant trait x treatment interaction after removing the batch effect & sample weight as a covariate. We computed a z-score of the $log_2$ abundance, where we adjusted the data, by metabolite, to have a mean of zero and a standard deviation of 1. The resulting heatmap presents the metabolites in rows and samples in columns, both of which are clustered. You can view the high resolution heatmap [here](`r paste0("plots/Heatmap_",datamode,"_treatment_trait_interactions.tiff")`) 


# Pathway analysis with Mummichog.

We used  *Mummichog* (version 1.0.10) [@Mummichog:2013], to perform putative metabolite annotation and pathway analysis.
The metabolic model Drosophila melanogaster (BioCyc 16.5) was used to test the enrichment of pathways and networks. All the metabolites that fitted in the univariate linear models were used as the input file (m/z values, retention time, FDR, and t-statistics), FDR cutoff was set at FDR of `r fdrcutoff` and 100 permutations were run to estimate the null distribution. Mode (-m) set as `r datamode`. Results are included in the mummichog_output subfolder. 



# R packages used
  
  
```{r, echo = FALSE}

sessionInfo()

```

# References




